Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-24 RNARedPrint.tex
add R-squared

Fig.~\ref{fig:energydist-pk} illustrates how well complex realistic energy models can be approximated based on simpler, but better tractable ones. For
the two target structures of Fig.~\ref{fig:energydist-pk}B,
Fig.~\ref{fig:energydist-pk}A shows the good fit between realistic
energies in the full-fledged Dirks and Pierce energy model for
pseudoknots (D\&P model) and energies in the stacking energy
model. For the shown fits we sampled $n=1\,000$ sequences, targeting a
\GCb\% of $60\%$.
%
For an example instance of the \texttt{Modena} benchmark with two
pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
the Turner energy distributions of the single structures as they
result from sampling with different weight parameters. The figure
illustrates how our multidimensional Boltzmann sampling strategy can,
to a large extent, independently shift the Turner energies of sampled
sequences towards prescribed targets. See Supp.\ Mat.~\ref{appsec:illustrating-mdbs} for a further example with three
pseudoknot-free target structures.}

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.97\textwidth]{Figs/energy_shift}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling. \textbf{(A)} Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for sampled sequences; the good
    match enables more efficient targeting of Turner energies based on
    targeting stacking model energies.  We show the fits for both
    target structures R1 and R2 of subfigure B.  \textbf{(B)} D\&P energy
    distributions for two target structures R1 and R2. We show the
    results of targeting the respective free energies $(-30,-20)$,
    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
    structures---this demonstrates the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure; moreover, for
    comparison, the distributions for uniform and Boltzmann
    samples---respectively associated with homogenous weights $1$ and
    $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}

@@

The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
Boltzmann-sampled sequences outperform uniform seeds on every dataset,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is be observed for both of the terms in \Obj{} (see
Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
produces sequences whose targets are both substantially closer to
the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{appsec:detailed-benchmark-results}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.

%
Moreover, for all instances as well, the superiority of Boltzmann designs
remains unchallenged even after local optimizations, as shown by
Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{appsec:contributions-multidefect}
and \ref{appsec:detailed-benchmark-results}. This observation is consistent with a
superior quality of the starting point for the greedy walk, probably leading
to better local minima of the \Obj{} function. However, it should be noted
that the greedy walk is based on the uniform sampling of
\RNAblueprint{}, and thus can be expected to partially level the
advantages of Boltzmann sampling. In future work, we hope to improve
this aspect by exploiting Boltzmann sampling during the
optimization run.

@@

We introduced a novel approach, based on a general framework and efficient algorithms, to design RNA
sequences, while targeting very specific complex properties. In particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
Our method combines an FPT stochastic sampling algorithm
with multi-dimensional Boltzmann sampling over distributions
controlled by expressive RNA energy models. Compared to the
previously best available sampling method (uniform sampling), this
approach generated significantly better seed sequences for instances
of an extensive multi-target design benchmark, including pseudoknots.
%
To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling which,
from a complexity-theoretic point of view, motivates the FPT, tree decomposition-based nature of our method.

>>>>

Fig.~\ref{fig:energydist-pk} illustrates how well complex realistic energy models can be approximated based on simpler, but better tractable ones. For
the two target structures of Fig.~\ref{fig:energydist-pk}B,
Fig.~\ref{fig:energydist-pk}A shows the good fit between realistic
energies in the full-fledged Dirks and Pierce energy model for
pseudoknots (D\&P model) and energies in the stacking energy
model\revised{, which is obtained for each of the two target structures (with respective $R^2$ values of $0.846$ and $0.841$)}. For the shown fits we sampled $n=1\,000$ sequences, targeting a
\GCb\% of $60\%$.
%
For an example instance of the \texttt{Modena} benchmark with two
pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
the Turner energy distributions of the single structures as they
result from sampling with different weight parameters. The figure
illustrates how our multidimensional Boltzmann sampling strategy can,
to a large extent, independently shift the Turner energies of sampled
sequences towards prescribed targets. See Supp.\ Mat.~\ref{appsec:illustrating-mdbs} for a further example with three
pseudoknot-free target structures.}

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.97\textwidth]{Figs/energy_shift}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling. \textbf{(A)} Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for the sampled sequences and both target structures R1 and R2. The good
    match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
    targeting stacking model energies.  We show the fits for both
    target structures R1 and R2 of subfigure B.  \textbf{(B)} D\&P energy
    distributions for two target structures R1 and R2. We show the
    results of targeting the respective free energies $(-30,-20)$,
    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
    structures---this demonstrates the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure; moreover, for
    comparison, the distributions for uniform and Boltzmann
    samples---respectively associated with homogenous weights $1$ and
    $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}

@@

The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
Boltzmann-sampled sequences outperform uniform seeds on every dataset,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is observed for both of the terms in \Obj{} (see
Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
produces sequences whose targets are both substantially closer to
the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{appsec:detailed-benchmark-results}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.

%
Moreover, for all instances as well, the Boltzmann designs
remain superior even after local optimizations, as shown by
Fig.~\ref{fig:benchmark-results} and Supp. Mat.~\ref{appsec:contributions-multidefect}
and \ref{appsec:detailed-benchmark-results}. This observation is consistent with a
superior quality of the starting point for the greedy walk, probably leading
to better local minima of the \Obj{} function. However, it should be noted
that the greedy walk is based on the uniform sampling of
\RNAblueprint{}, and thus can be expected to partially level the
advantages of Boltzmann sampling. In future work, we hope to improve
this aspect by exploiting Boltzmann sampling during the
optimization run.

@@

We introduced a novel approach, based on a general framework and efficient algorithms, to design RNA
sequences, while targeting very specific complex properties. In particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
Our method combines an FPT stochastic sampling algorithm
with multi-dimensional Boltzmann sampling over distributions
controlled by expressive RNA energy models. \revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice, which is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework. In a proof-of-concept study 
  on an established benchmark set for multi-target RNA design (including pseudoknotted instances), 
  the approach generated significantly better seed sequences than
the previously best available method (uniform sampling). These results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.}
%
To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling which,
from a complexity-theoretic point of view, motivates the FPT, tree decomposition-based nature of our method.
